Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-01-28 papers/wdmerger_I/paper.tex
Add description to the paper of the rotation coupling to the hydrodynamics.

For the evolution of binary systems, it is most natural to evolve the
two stars in a frame that is co-rotating at the same period as the
orbital period. CASTRO has the ability to evolve systems in a rotating
reference frame, as discussed in the original code paper. Source terms
corresponding to the Coriolis and centrifugal force terms are added to
the momentum and energy equations. In this frame, the stars
essentially remain stationary in their original positions due to the
centrifugal force supporting against the gravitational attraction, and
will remain this way as long as significant mass transfer does not
occur. \cite{swc:2000} demonstrated (in the context of neutron star
mergers) that conservation of angular momentum is much easier to
obtain in the rotating reference frame than in an inertial frame in
which stars advect large amounts of material around the domain. We
wish to emphasize that although it is commonly stated in the
literature that grid-based codes poorly conserve angular momentum,
this is not generally true. When the resolution is sufficiently high
and a rotating reference frame is employed, excellent conservation
properties can result, as demonstrated in Section \ref{sec:kepler}. We
note that as the stars begin to coalesce, the rotating reference frame
will no longer provide a good approximation to the spatial motion of
the stars and then they will begin to significantly move around the
domain. This is not necessarily problematic because the most important
feature of the rotating frame is that it helps ensure that the initial
coalescence is not the result of spurious numerical loss of angular
momentum. When significant mass transfer sets in and evolution
proceeds on a dynamical timescale, the conservation properties may be
slightly worse but angular momentum conservation is also less
important.

@@

We use the Castro code as described in \citet{castro}.  For all the
runs, the PPM reconstruction is done, using the original limiters for
the parabolic profiles \citep{ppm}.  The prediction of the interface
states appears as:
\begin{equation}
\label{eq:ppmstatel}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L -
   \sum_{\nu;\lambda_i^{(\nu)}\ge 0} l_i^{(\nu)} \cdot \left [
        \tilde{q}_L  - \mathcal{I}^{(\nu)}_+(q_i)
       \right ] r_i^{(\nu)}
\end{equation}
where $q$ is the vector of primitive variables, $l_i^{(\nu)}$ and
$r_i^{(\nu)}$ are the left and right eigenvectors with eigenvalue
$\lambda_i^{(\nu)}$, with $\nu$ the index of the characteristic wave of
the system.  The sum is over all the waves that result from the
characteristic structure of the problem, but designed such that only
waves moving toward the interface contribute to the interface value,
$q_{i+1/2,L}^{n+1/2}$.  The reference state, $\tilde{q}_L$ is
chosen to minimize the work of the characteristic projection.
Finally, $\mathcal{I}_+^{(\nu)}(q)$ is the
average under the parabolic profile of quantity $q$ of all the
information that can reach the right interface of the zone $i$ as
carried by the wave $\nu$.  The reader is referred to
\citet{ppmunsplit} for further defaults.

@@

\subsection{Gravity Tracing}

We note a few additional differences between the original PPM
implementation of \citet{ppm} and Castro.  In the original PPM
implementation, the gravitational was reconstructed as a parabola, and
this was traced under to find the forcing that affects the interface
for each wave.  Castro followed \citet{ppmunsplit} which instead adds
$(\Delta t/2)g$ to the interface states at the end of the
reconstruction.  In the current implementation, we do the original
parabolic reconstruction and characteristic tracing of the gravitation
source.  Our system with the source appears as:
\begin{equation}
q_t + A(q) q_x = G
\end{equation}
where $G = (0, g, 0)^T$---i.e. the gravitational source only affects
$u$, not $\rho$ or $p$.  Note that in the PPM paper, they put $G$ on 
the lefthand side of the primitive variable equation, so our signs are
opposite.  Our projections are now:
\begin{equation}
\sum_{\nu; \lambda^\enu \ge 0}l^\enu \cdot (\tilde{q} - \mathcal{I}^\enu_+(q) - \tfrac{\Delta t}{2} G) r^\enu
\end{equation}
for the left state, and
\begin{equation}
\sum_{\nu; \lambda^\enu \le 0} l^\enu \cdot (\tilde{q} - \mathcal{I}^\enu_-(q) - \tfrac{\Delta t}{2} G) r^\enu 
\end{equation}
for the right state.  Since $G$ is only non-zero for velocity, only
the velocity changes.  Writing out the sum (and performing the vector products), we
get:
\begin{eqnarray}
u_{i+1/2,L}^{n+1/2} =
   \tilde{u}_+ 
  &-& \frac{1}{2} \left [
      \left (\tilde{u}_+ - \mathcal{I}_+^\evm(u) - \frac{\Delta t}{2} \mathcal{I}^\evm_+(g) \right ) - 
       \frac{\tilde{p}_+ - \mathcal{I}_+^\evm(p)}{C} \right ] \nonumber \\
  &-& \frac{1}{2} \left [
      \left (\tilde{u}_+ - \mathcal{I}_+^\evp(u) - \frac{\Delta t}{2} \mathcal{I}^\evp_+(g) \right ) +
       \frac{\tilde{p}_+ - \mathcal{I}_+^\evp(p)}{C} \right ]
\end{eqnarray}
These differ from the expression in the PPM paper, where $\Delta t G$,
not $\Delta t/2 G$ is used in the projection, however we believe that
the factor of $1/2$ is correct.  To see this, notice that if both
waves are moving toward the interface, then the source term that is
added to the interface state is $(\Delta t/4) (\mathcal{I}_+^\evm(g) +
\mathcal{I}_+^\evp(g))$ for the left state, which reduces to $(\Delta
t/2) g$ for constant g---this matches the result from Taylor
expanding to the interface at the half-time (as in \citealt{ppmunsplit}).

@@

The tracing of gravity can be controlled in the public version of Castro
with the parameter {\tt castro.ppm\_trace\_grav}.

>>>>

For the evolution of binary systems, it is most natural to evolve the
two stars in a frame that is co-rotating at the same period as the
orbital period. Since the publication of the original code paper, CASTRO 
now has the ability to evolve systems in a rotating reference frame. 
Source terms corresponding to the Coriolis and centrifugal 
force terms are added to the momentum and energy equations. In this frame, 
the stars essentially remain stationary in their original positions due to the
centrifugal force supporting against the gravitational attraction, and
will remain this way as long as significant mass transfer does not
occur. \cite{swc:2000} demonstrated (in the context of neutron star
mergers) that conservation of angular momentum is much easier to
obtain in the rotating reference frame than in an inertial frame in
which stars advect large amounts of material around the domain. We
wish to emphasize that although it is commonly stated in the
literature that grid-based codes poorly conserve angular momentum,
it is only generally true that grid-based codes do not exactly conserve 
angular momentum when the equations are written in conservative form
for linear momentum. (See also \cite{motl:2002} for an example of how 
to evolve the hydrodynamics equations for angular momentum.) 
The extent to which angular momentum conservation is violated
will be a function of the resolution. When this is sufficiently high
and a rotating reference frame is employed, excellent conservation
properties can result, as demonstrated in Section \ref{sec:kepler}. 
In particular, at reasonable resolution for a binary orbit our code 
conserves angular momentum well enough that this no longer becomes 
a reason to use a Lagrangian method such as smoothed-particle hydrodynamics.
We note that as the stars begin to coalesce, the rotating reference frame
will no longer provide a good approximation to the spatial motion of
the stars and then they will begin to significantly move around the
domain. This is not necessarily problematic because the most important
feature of the rotating frame is that it helps ensure that the initial
coalescence is not the result of spurious numerical loss of angular
momentum. When significant mass transfer sets in and evolution
proceeds on a dynamical timescale, the conservation properties may be
slightly worse but angular momentum conservation is also less
important.

In a rotating reference frame with angular frequency vector $\bm{\omega}$, the momentum equation becomes:
\begin{equation}
  \left.\frac{\partial(\rho \mathbf{u})}{\partial t}\right|_{\text{rot}} = -2\, {\bm\omega} \times (\rho\mathbf{u}) - \rho {\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right).
\end{equation}
Here $\mathbf{r}$ is the position vector with respect to the origin. Typically we choose $\bm{\omega} = (0, 0, 2\pi / T)^T$,
with the rotation axis coincident with the $z$ axis at $x = y = 0$.
$T$ is the rotation period, which is the most natural quantity to specify
for a rotating stellar system. As described in the appendix, we include this source term
in the edge state prediction in a way that is analogous to the gravity source.
We evaluate all quantities at cell centers and use the same predictor-corrector 
approach that we use for the gravity source terms to the momentum equations.

The update to the energy equation can be determined by taking the dot product of the velocity
with the momentum source terms. The Coriolis term vanishes identically, as it must
because the Coriolis term does no work on the fluid. The update from the centrifugal force becomes
\begin{equation}
  \left.\frac{\partial(\rho E)}{\partial t}\right|_{\text{rot}} = \frac{1}{\Delta V}\int \rho \mathbf{u} \cdot \mathbf{f}_R\, dV,
\end{equation}
with $\mathbf{f}_R \equiv  -{\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right)$. 
This expression is identical to the gravity source under the interchange of $\mathbf{g}$ with $\mathbf{f}_R$.
As observed by \cite{marcello:2012}, we can similarly write down a rotational potential,
\begin{equation}
  \Phi_R = -\frac{1}{2} \left| {\bm\omega} \times \mathbf{r} \right|^2.
\end{equation}
Therefore we can retrace the steps of Section \ref{sec:gravity_hydro_coupling} to obtain
an update to $(\rho E)$ that conserves the total energy in the presence of rotation,
\begin{equation}
  (\rho E_{\text{tot}}) = (\rho E) + \frac{1}{2} \rho \Phi + \rho \Phi_R.
\end{equation}
In one dimension, this would be
\begin{align}
  \left.\Delta(\rho E_i)\right|_{\text{rot}} &= \frac{1}{2} \Delta m_{i-1/2} (\Phi_{R_{i}} - \Phi_{R_{i-1}}) \notag \\
  &+ \frac{1}{2} \Delta m_{i+1/2} (\Phi_{R_{i+1}} - \Phi_{R_{i}}).
\end{align}
As with gravity, in three dimensions we also include the mass fluxes through all six cell faces.

@@

For all the runs, the PPM reconstruction is done using the original limiters for
the parabolic profiles \citep{ppm}.  The prediction of the interface
states appears as:
\begin{equation}
\label{eq:ppmstatel}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L -
   \sum_{\nu;\lambda_i^{(\nu)}\ge 0} l_i^{(\nu)} \cdot \left [
        \tilde{q}_L  - \mathcal{I}^{(\nu)}_+(q_i)
       \right ] r_i^{(\nu)}
\end{equation}
where $q$ is the vector of primitive variables, $l_i^{(\nu)}$ and
$r_i^{(\nu)}$ are the left and right eigenvectors with eigenvalue
$\lambda_i^{(\nu)}$, with $\nu$ the index of the characteristic wave of
the system.  The sum is over all the waves that result from the
characteristic structure of the problem, but designed such that only
waves moving toward the interface contribute to the interface value,
$q_{i+1/2,L}^{n+1/2}$.  The reference state, $\tilde{q}_L$ is
chosen to minimize the work of the characteristic projection.
Finally, $\mathcal{I}_+^{(\nu)}(q)$ is the
average under the parabolic profile of quantity $q$ of all the
information that can reach the right interface of the zone $i$ as
carried by the wave $\nu$.  The reader is referred to
\citet{ppmunsplit} for further defaults.

@@

\subsection{Gravity and Rotation Tracing}

We note a few additional differences between the original PPM
implementation of \citet{ppm} and CASTRO.  In the original PPM
implementation, the gravitational acceleration was reconstructed as a parabola, and
this was traced under to find the forcing that affects the interface
for each wave.  CASTRO followed \citet{ppmunsplit} which instead adds
$(\Delta t/2)g$ to the interface states at the end of the
reconstruction.  In the current implementation, we do the original
parabolic reconstruction and characteristic tracing of the gravitation
source.  Our system with the source appears as:
\begin{equation}
q_t + A(q) q_x = G
\end{equation}
where $G = (0, g, 0)^T$---i.e. the gravitational source only affects
$u$, not $\rho$ or $p$.  Note that in the PPM paper, they put $G$ on 
the lefthand side of the primitive variable equation, so our signs are
opposite.  Our projections are now:
\begin{equation}
\sum_{\nu; \lambda^\enu \ge 0}l^\enu \cdot (\tilde{q} - \mathcal{I}^\enu_+(q) - \tfrac{\Delta t}{2} G) r^\enu
\end{equation}
for the left state, and
\begin{equation}
\sum_{\nu; \lambda^\enu \le 0} l^\enu \cdot (\tilde{q} - \mathcal{I}^\enu_-(q) - \tfrac{\Delta t}{2} G) r^\enu 
\end{equation}
for the right state.  Since $G$ is only non-zero for velocity, only
the velocity changes.  Writing out the sum (and performing the vector products), we
get:
\begin{eqnarray}
u_{i+1/2,L}^{n+1/2} =
   \tilde{u}_+ 
  &-& \frac{1}{2} \left [
      \left (\tilde{u}_+ - \mathcal{I}_+^\evm(u) - \frac{\Delta t}{2} \mathcal{I}^\evm_+(g) \right ) - 
       \frac{\tilde{p}_+ - \mathcal{I}_+^\evm(p)}{C} \right ] \nonumber \\
  &-& \frac{1}{2} \left [
      \left (\tilde{u}_+ - \mathcal{I}_+^\evp(u) - \frac{\Delta t}{2} \mathcal{I}^\evp_+(g) \right ) +
       \frac{\tilde{p}_+ - \mathcal{I}_+^\evp(p)}{C} \right ]
\end{eqnarray}
These differ from the expression in the PPM paper, where $\Delta t G$,
not $\Delta t/2 G$ is used in the projection, however we believe that
the factor of $1/2$ is correct.  To see this, notice that if both
waves are moving toward the interface, then the source term that is
added to the interface state is $(\Delta t/4) (\mathcal{I}_+^\evm(g) +
\mathcal{I}_+^\evp(g))$ for the left state, which reduces to $(\Delta
t/2) g$ for constant g---this matches the result from Taylor
expanding to the interface at the half-time (as in \citealt{ppmunsplit}).

@@

The tracing of gravity can be controlled in the public version of CASTRO
with the parameter {\tt castro.ppm\_trace\_grav}.

Similarly, we also trace under the centrifugal and Coriolis forces to capture
the effects of rotation on the edge states. For the primitive variables, the rotation
source term is $F_R = (0, f_R, 0)^T$, where $f_R = -2\, {\bm{\omega}} \times \mathbf{u} - {\bm\omega}\times \left({\bm\omega} \times \mathbf{r}\right).$
Just like the gravity source, the rotation source terms only update the velocities. Therefore 
the rotation source terms can be included in a manner identical to that of the gravity source
terms, with the replacement of $G$ with $F_R$ and $g$ with $f_R$ in the above. This is controlled in CASTRO
with the parameter {\tt castro.ppm\_trace\_rot}.
